On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-09 paper/application.tex
update, rough draft

\subsection{3D PPD}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). At each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux, given
the $\alpha(R)$ and $Q(R)$ profiles obtained from the 2D model
above. We then solve the 3D eigenvalue problem as in
\S\ref{3ddisk}, with the additional boundary condition for the temperature 
perturbation $\delta T^\prime(0) = \delta T(\zmax) = 0$.

We use the same disk model as in \S\ref{pp2d_example} but without
irradiation ($T_\mathrm{irr}=0$), and consider a smaller domain
with $R\in[10,100]AU$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous,
    self-gravitational modes.  
    \label{rafikov_growth3d}}
\end{figure}

>>>>

\subsection{3D PPD}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
from the 2D model above, at each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
We use the same disk model as in \S\ref{pp2d_example} but without 
irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density

Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
wavenumber for $R\in[10,100]$AU. We also plot the corresponding 2D
results with softened self-gravity. There is a close match for
$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
1$) and both models apply. However, beyond $60$AU where the disk is
optically thin, radiative diffusion (the 3D curve) does not apply and
under-estimates the growth rates. Nevertheless, the comparison shows
that the transition radius $R\sim60$AU can be calculated within the 2D
framework. This is not surprising since the instability is 
fundamentally driven by viscous and self-gravity forces in the
momentum equations, not the energy equation.

%the instability is due to viscosity/gravity in momenta equation so
%details of energy eq doesn't matter

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous, 
    self-gravitating unstable modes. The black curves are obtained
    from the 3D eigenvalue problem with radiative diffusion, while the
    green curves are 
    \label{rafikov_growth3d}}
\end{figure}

### 2016-01-09 paper/paper3d.tex
update, rough draft

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{TBD}
\shortauthors{M-K.\ Lin, K. Kratter}

@@

%\title{untitled linear problem}
\title{On the Gravitational stability of gravito-turbulent disks}
\author{Min-Kai Lin\altaffilmark{1} \& Kaitlin M. Kratter}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
stuff
%on the gravitational stability of gravito-turbulent disks
\end{abstract}

\section{Introduction}\label{intro}
%normal GI and viscous GI. appplicability of simple stability criteria
%and dispersion relation (toomre and cooling) to complex
%systems.

%non-existence of gravito-turbulent state for fast cooling
%because such states are dynamically unstable. secondary gi in two
%ways: classic, pure gravity; viscosity-assisted.

%provide generalized dispersion relations.

%3D calculations.

%apply linear theory
%to real disk models.

%notion of max alpha comes from simulations. 
%max alpha criteria applies in local simulations, so not to do with how
%fast gi can transport material in a global disk.

%combine independent fragmentation conditions Q less than one and beta
%less than a few in a single framework: these determine growth rate of
%viscous GI.

%toomre Q and standard dispersion relation here

%real disks have cooling (removes pressure support)
%and turbulence (removes rotational support ---  if it's like a
%viscosity)

%indeed fragmenting disks have cooling and turbulence

%so discussion of GI in real disks should have both

Understanding the gravitational stability (GI) of rotating disks   
is fundamental to many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. First, it can
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards. Secondly, it may lead to disk
fragmentation, which has been invoked to explain planet formation at
large radii. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, significant progress can be
made through analytical understanding of GI.  %physical insight

It is now customary to begin any discussion of disk GI with 
the \emph{Toomre criterion}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}<1 
\end{align}
for instability \citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic 
frequency (which equals the rotation frequency $\Omega$ in a Keplerian
disk), $\Sigma$ is the surface density and $G$ is the
gravitational constant. This condition describes how self-gravity
($\Sigma$) destabilizes the disk against 
rotation ($\kappa$) and pressure ($c_s$). Eq. \ref{toomreQ_criterion}
formally applies to local, axisymmetric disturbances in a
two-dimensional (2D, razor-thin), inviscid and adiabatic (including
isothermal) disk.

@@

Despite being a result from ideal disk models, the Toomre $Q$  
parameter is widely applied to assess GI in more general 
situtations. This includes non-ideal, realistic disk structures
obtained with additional physics not encapsulated by $Q$
\citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
leave the stability condition unaltered. Indeed, we show that GI in more
realistic disk models cannot be described by a single `Toomre-like' 
condition.

@@

The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in 
realistic PPD models. Specifically, we will examine the effect of
cooling and viscosity on GI. Their relevance to PPDs are described
below.

@@

\subsection{Cooling removes pressure support}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature is relaxed
to some pre-defined value, which may be zero. It is 
often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time.

{\bf physical origin of cooling in PPDs}

The Toomre $Q$ parameter does not include cooling, and thus cannot be 
used to assess GI in realistic disks. Indeed, there exists a separate, 
\emph{empirical} cooling time criterion which governs stability against disk
fragmentation (see below), but its precise value is uncertain.

We shall quantify, using linear stability analyses, how cooling 
destablize disks. We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

\subsection{Viscosity removes rotational support}
It is well known that viscous, self-gravitating disks can develop 
GI even when $Q>1$. This is because viscosity compromises rotational
support against self-gravity for long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. %also in dusty fluids -
                                %youdin 11 and the inutsuka paper

Now, accretion disks are often modeled with a Navier-Stokes viscosity
to mimic hydrodynamic or magneto-hydrodynamic turbulence
\citep{shakura73}. This involves writing the
kinematic viscosity $\nu$ as  
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form.

How well turbulence can be modeled as an effective, Navier-Stokes
viscosity is a separate issue \citep{balbus99}. However, it is 
reasonable to assume that turbulence frustates rotational 
support. (This is, after all, a physical cause for mass accretion.)

@@

\subsection{Relation to gravito-turbulent disk fragmentation}%and ring
                                %formation 
%can the continous formation/destruction of clumps (as seen in sims)
%be interpreted as `overstability'? 
%for overstable modes over-density oscilates 
The two effects described above are immediately relevant to
`gravito-turbulence'. This is a steady disk state in which cooling is
balanced by turbulent heating, itself caused by GI. In this case 
the $\alpha$ and $\beta$ parameters are inversely related.

Experimentally, however, it is found that if the imposed $\beta$ is
too small, 
\begin{align}
  \beta < \beta_\mathrm{c},
\end{align}
then the disk fragments. There is considerable debate on the exact
value of $\beta_\mathrm{c}$ and whether or not a critical cooling time can
be defined at all. However, it is generally accepted that \emph{
  steady, gravito-turbulent disks do not exist for sufficiently rapid
cooling}.

We do not attempt to model disk fragmentation from first
principles. Instead, we simply ask: \emph{when does a 
  gravito-turbulent disk become dynamically unstable?} 
%The two destabilizing agents are cooling and viscosity  
We will find the answer to be consistent with the current consensus on
disk fragmentation.

@@

\subsection{Plan of this paper}
%plan of this paper 
%aim to analyse GI in realistic disks with minimal input from numerical
%simulation

>>>>

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Non-ideal GI}
\shortauthors{M-K.\ Lin, K. Kratter}

@@

%\title{untitled linear problem}
\title{On the Gravitational stability of gravito-turbulent accretion disks}
\author{Min-Kai Lin\altaffilmark{1} \& Kaitlin M. Kratter}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
Low mass, self-gravitating accretion disks are able to sustain a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues.   
We explore the physical interpretation of this result as the
compromise of gravitational stability by non-ideal effects: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
We derive a flexible dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. 
Thus, a steady gravito-turbulent state cannot persist for more and a
few orbits in the outer disk, coincident with the expected region for
fragmentation based on emperical conditions inferred from numerical
simulations. 
%This is understood within our framework as the disk's
%inability to sustain too large of a stress as it undermines
%rotational support against self-gravity. 
%This motivates an interpretation of disk 
%fragmentation as a secondary gravitational instability enabled by 
%the underlying gravito-turbulent viscosity when it exceeds a critical value.         
\end{abstract}
%opacity

\section{Introduction}\label{intro}
Understanding the gravitational stability (GI) of rotating disks   
is fundamental to many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. It can 
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
large radii \citep{helled14}. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

The standard metric for the (inverse) strength of disk self-gravity is the 
\emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure
($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
axisymmetric disturbances in a two-dimensional (2D, razor-thin),
inviscid and adiabatic (including isothermal) disk.

@@

The Toomre  parameter is widely applied to assess GI in more general 
situtations (cf. the ideal disk models from which $Q$ is derived
from).  This includes realistic disk structures
obtained with non-ideal physics that are not explicitly contained in
$Q$ \citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
leave the stability condition unaltered. Indeed, we show that GI in more
realistic disk models cannot be described by a single `Toomre-like' 
condition.

@@

The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in realistic accretion
disk models. Specifically, we will examine the effect of cooling and
viscosity on GI in PPDs.  
%Their relevance to PPDs are described below.

@@

\subsection{Cooling removes pressure support}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
-(T-T_0)/\tcool$. It is often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of cooling function, first applied by \cite{gammie01}, is a
simple parameterization that allows the full range of thermodynamic
response to be explored. 
{\bf physical origin of cooling in PPDs}

The Toomre $Q$ parameter does not include cooling, and thus cannot be
formally applied to assess GI in realistic disks. Indeed, there exists
a separate, \emph{empirical} cooling time criterion which governs
stability against disk fragmentation (see below), but this is not well
understood.

We shall quantify how cooling removes pressure
support to enable GI even when $Q>1$. 
We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

\subsection{Viscosity removes rotational support}
It is well known that viscous, self-gravitating disks can develop 
GI even when $Q>1$. This is because viscosity compromises rotational
support against self-gravity for long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect 
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag forces \citep{youdin11,takahashi14}.

Now, gaseous accretion disks are often modeled with a Navier-Stokes viscosity
to mimic hydrodynamic or magneto-hydrodynamic turbulence
\citep{shakura73}. This involves writing the
kinematic viscosity $\nu$ as  
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form.

How well turbulence can be modeled as an effective, Navier-Stokes
viscosity is a separate issue \citep{balbus99}. However, it is 
reasonable to assume that turbulence frustates rotational 
support. (This is, after all, the physical cause for turbulent mass
accretion.)

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}%and ring
                                %formation 
%can the continous formation/destruction of clumps (as seen in sims)
%be interpreted as `overstability'? 
%for overstable modes over-density oscilates 
The two effects described above are immediately relevant to
`gravito-turbulence'. This is a steady disk state in which cooling is
balanced by turbulent heating, itself caused by GI. In this case 
the $\alpha$ and $\beta$ parameters are inversely related
\citep{gammie01}.

Experimentally, however, it is found that if $\beta$ ($\alpha$) is
too small (large), 
\begin{align}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. 
There is considerable debate on the exact value of $\beta_\mathrm{c}$
and whether or not a critical cooling time can 
be defined at all \citep{meru11,meru12,paardekooper12}. However, it is
generally accepted that \emph{steady, gravito-turbulent disks do not
  exist for sufficiently rapid cooling}.

The non-existence of steady gravito-turbulence for rapid cooling, or
equivalently large viscosity, is intriguing because, as discussed
above, these effects reduce pressure and rotational support 
against self-gravity, respectively.

In this paper, we consider the stability of PPDs subject to 
gravito-turbulence modeled as an effective viscosity. 
We perform standard stability analyses with explicit inclusion of the
relevant physics. 
We will find that the
regions of a gravito-turbulent PPD that is dynamically unstable,
agrees with the current consensus on where, and the conditions under
which, a PPD is expected to fragment. This motivates an alternative 
interpretation of disk fragmentation as the inability to maintain a
stable, gravito-turbulent state due to the action of cooling and
viscosity.

%We do not attempt to model disk fragmentation from first
%principles. Instead, we simply ask: \emph{when does a 
%  gravito-turbulent disk become dynamically unstable?} 
%We will find the answer to be consistent with the current consensus on
%disk fragmentation.

@@

\subsection{Plan}
%plan of this paper 
%aim to analyse GI in realistic disks with minimal input from numerical
%simulation
